---
title: "Women's Soccer Data"
output: html_notebook
---

We'll use this notebook to explore some NCAA women's soccer data and [our questions](https://docs.google.com/spreadsheets/d/1kTSrnl1URmcexIWLDMWqDnp3q-c2OS45uifnhjp10n4/edit#gid=0). First, let's load the tidyverse library:

```{r}
library(tidyverse)
```

We'll want to start with 2022 match data, which is located in [this GitHub repository](https://github.com/dwillis/NCAAWomensSoccerData). We can load the match data directly from GitHub using `read_csv`:


```{r}
matches_2022 <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/main/data/ncaa_womens_soccer_matchstats_2022.csv")
```

Let's take a look at this data to see what it contains, using `head` and `glimpse`:

```{r}
head(matches_2022)
glimpse(matches_2022)
```

Now, onto the questions using aggregates, which means `group_by` and `summarize`:

1. How many shots in total has Maryland attempted and how many have been on goal? Calculate a percentage of shots on goal.

```{r}
matches_2022 %>% 
  group_by(team) %>%
    summarize(total_shots = sum(sh_att), sog = sum(so_g)) %>%
      mutate(percentage_quality_shots = 100*sog/total_shots) %>%
  arrange(desc(percentage_quality_shots))

 
```

2. What about the same shot stats from last season's first five games? You'll need to load the matches from 2021 - just change the season.

```{r}
matches_2021 <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/main/data/ncaa_womens_soccer_matchstats_2021.csv")
matches_2021 %>% 
  group_by(team) %>%
  filter(team == "Maryland Terrapins, Terps" & date < "2021-09-05") %>%
    summarize(total_shots = sum(sh_att), sog = sum(so_g)) %>%
      mutate(percentage_quality_shots = 100*sog/total_shots) %>%
  arrange(desc(percentage_quality_shots))

  
```


3. What about draws - Maryland started with 4 ties. Have any other teams had as many?

```{r}
matches_2022 %>% 
    group_by(team) %>%
  filter(outcome == "Draw") %>%
    summarize(draws = n()) %>%
  arrange(desc(draws))

```


4. How many total corner kicks has Maryland had so far and how many have its opponents have? Calculate the difference between the two totals for every team.

```{r}
matches_2022 %>% 
  filter(team == "Maryland Terrapins, Terps") %>%
    summarize(date, team, opponent, result, corners, defensive_corners) %>%
      mutate(corner_diff = corners - defensive_corners) %>% 
  arrange(desc(corner_diff))
```

```{r}
correlations <- read_csv("C:/Users/jetso/Documents/jour479x_fall_2022/data/ncaa_womens_soccer_matchstats_2022.csv")
```
```{r}
newcorrelations <- correlations %>% 
  mutate(
    differential = goals - defensive_goals
    )
```


```{r}
newcorrelations %>% summarise(correlation = cor(differential, fouls, method="pearson"))

```
```{r}
newcorrelations %>% 
  summarise(correlation = cor(differential, yellow_cards, method="pearson"))
```
```{r}
fit <- lm(differential ~ fouls, data = newcorrelations)
summary(fit)
```

